Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2024-07-25 manuscript/appendix.qmd
Final edits pre-II

---
title: 'Online appendix for "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"'
short-title: 'Online appendix for "Pandemic Pass?"'
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: July 22, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
crossref: 
  tbl-title: "Table&nbsp;A"
  fig-title: "Figure&nbsp;A"
---

@@

Complete results from all the models, along with posterior predictive checks, goodness-of-fit measures, and prediction diagnostics are all available at a companion statistical analysis compendium at <https://doi.org/10.17605/OSF.IO/ANONYMIZED-FOR-NOW>.

@@

```{r tbl-results-determinants}
#| tbl-cap: Results from Authors (2024) showing the determinants of derogation
#| include: true
#| echo: false

>>>>

---
title: 'Online appendix for "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"'
short-title: 'Online appendix for "Pandemic Pass?"'
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: July 25, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
crossref: 
  tbl-title: "Table&nbsp;A"
  fig-title: "Figure&nbsp;A"
---

@@

Complete results from all the models, along with posterior predictive checks, goodness-of-fit measures, and prediction diagnostics are all available at a companion statistical analysis compendium at <https://doi.org/10.17605/OSF.IO/7GVDA>.

@@

```{r tbl-results-determinants}
#| tbl-cap: Results from @ChaudhryComstockHeiss:2024 showing the determinants of derogation
#| include: true
#| echo: false

### 2024-07-25 manuscript/manuscript.qmd
Use references div
Final edits pre-II

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  This research note asks whether states issuing pandemic-era human rights treaty derogations implemented emergency provisions as intended or used them to abuse human rights during a time of crisis. In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states that derogated from their international human rights obligations imposed emergency measures that were temporary and did not violate non-derogable rights. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law and show that states are being sincere about their intentions and not, generally, using these mechanisms to cover abusive behavior.
thanks: |
  We would like to thank particpants of the 2022 Peace Science Society conference and the 2023 International Studies Association (ISA) convention for their comments and feedback.
date: July 22, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
---

@@

[^future]: This project specifically examines what happens after derogations are filed. In a separate project by Authors (2024) we explore the determinants of derogations and find that (1) pandemic-era backsliding and (2) the severity of the pandemic as measured by national-level deaths are significant and positive indicators of pandemic-related derogation submission, and that a country's level of rule of law is positively but not significantly predictive of derogation (see Table A7 in the appendix).

@@

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  This research note asks whether states issuing pandemic-era human rights treaty derogations implemented emergency provisions as intended or used them to abuse human rights during a time of crisis. In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states that derogated from their international human rights obligations imposed emergency measures that were temporary and did not violate non-derogable rights. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law and show that states are being sincere about their intentions and not, generally, using these mechanisms to cover abusive behavior.
thanks: |
  We would like to thank particpants of the 2022 Peace Science Society conference and the 2023 International Studies Association (ISA) convention for their comments and feedback.
date: July 25, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
---

@@

[^future]: This project specifically examines what happens after derogations are filed. In a separate project by @ChaudhryComstockHeiss:2024 we explore the determinants of derogations and find that (1) pandemic-era backsliding and (2) the severity of the pandemic as measured by national-level deaths are significant and positive indicators of pandemic-related derogation submission, and that a country's level of rule of law is positively but not significantly predictive of derogation (see Table A7 in the appendix).

@@

::: {#refs}
:::
